---
title: "Exercises"
author: "Joddy"
date: "2023-10-09"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Strings

### Collapsing
1. In your own words, describe the difference between the sep and collapse arguments to str_c().




2. Use str_length() and str_sub() to extract the middle character from a string. What will you do if the string has an even number of characters?
```{r}

```

### Regular expressions
3. Given the corpus of common words in stringr::words, create regular expressions that find all words that:
```{r}
words <- stringr::words
```


4. Start with “y”.
```{r}

```

5. End with “x”
```{r}

```

6. Are exactly three letters long. (Don’t cheat by using str_length()!)
```{r}

```

7. Have seven letters or more.
```{r}

```

Since this list is long, you might want to use the match argument to str_view() to show only the matching or non-matching words.

8. Create regular expressions to find all words that:

Start with a vowel.

That only contain consonants. (Hint: thinking about matching “not”-vowels.)

End with ed, but not with eed.

End with ing or ise.

9. Create a regular expression that will match telephone numbers as commonly written in your country.

10. Describe the equivalents of ?, +, * in {m,n} form.

11. Create regular expressions to find all words that:

Start with three consonants.
Have three or more vowels in a row.
Have two or more vowel-consonant pairs in a row.

### Coding with stringr
12. For each of the following challenges, try solving it by using both a single regular expression, and a combination of multiple str_detect() calls.

Find all words that start or end with x.
```{r}

```

Find all words that start with a vowel and end with a consonant.
```{r}

```

13. From the Harvard sentences data, extract:

The first word from each sentence.
```{r}

```

All words ending in ing.
```{r}

```

All plurals.
```{r}

```

14. Split up a string like "apples, pears, and bananas" into individual components.
```{r}

```

Extra challenge (for after the workshop): Solve the beginner regexp crosswords at https://regexcrossword.com/challenges/beginner.

## Dates
1. What does the tzone argument to today() do? Why is it important?

2. Create a vector of dates giving the first day of every month in 2015. 
```{r}

```

3. Create a vector of dates giving the first day of every month in the current year.

```{r}

```

4. Create an interval of your lifespan so far using your birthday as the starting point, and today() as the end point, and then use years(1) and the floor() function to round your age to an integer.
```{r}

```

