---
title: "exercises"
author: "Joddy"
date: "2024-03-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Exercises

## Function fundamentals

### 1. It’s possible (although typically not useful) to call an anonymous function. Which of the two approaches below is correct? Why?

function(x) 3()

(function(x) 3)()

### 2. A good rule of thumb is that an anonymous function should fit on one line and shouldn’t need to use {}. Review your code. Where could you have used an anonymous function instead of a named function? Where should you have used a named function instead of an anonymous function?


### 3.

#### a. What are the three important components of a function? How do we find them using code?


#### b. Are there any exceptions to these components?


#### c. Find the components of sd().

```{r}

```

#### d. Find the components of length().

```{r}

```

#### e. How can we determine whether a function is a primitive without relying on body(), formals(), and environment()? Hint: search the documentation for "primitive" or enter ?primitive in the console.


#### f. Demonstrate the primitive function test with sd() and length().

```{r}

```

## Lexical scoping

### 1. What are the four principles that govern how R looks for values? Briefly describe each of them.


### 2. What does the following code return? Why? Describe how each of the three c’s is interpreted.

```{r}
c <- 10
c(c = c)
```

### 3. What does the following function return? Make a prediction before running the code yourself.

```{r}
f <- function(x) {
  f <- function(x) {
    f <- function() {
      x ^ 2
    }
    f() + 1
  }
  f(x) * 2
}
f(10)
```

## Writing functions

### 1. Write both_na(), a function that takes two vectors of the same length and returns the number of positions that have an NA in both vectors. Also include an example usage of it.

```{r}
both_na <- function() {
  
}

both_na()
```

### 2. Write the logic for a greeting function that says “good morning”, “good afternoon”, or “good evening”, depending on the time of day. Make sure it provides correct results for the three cases included.

```{r}
greet <- function(time = lubridate::now()) {
  hr <- lubridate::hour(time)
  
  # Write the function logic here
}
greet()

greet(ymd_h("2024-01-08:05"))

greet(ymd_h("2024-01-08:13"))

greet(ymd_h("2024-01-08:20"))
```