---
title: "Solutions"
author: "Joddy"
date: "2023-11-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Exercises

1. Using the mtcars dataset, compute the standard deviation of every column using the appropriate map function.

```{r}
map_dbl(mtcars, sd)
```

2. Using the penguins dataset, compute the standard deviation of every numeric column in the data frame.

```{r}
# If you don't have this package installed yet
install.packages("palmerpenguins")
library(palmerpenguins)

penguins <- palmerpenguins::penguins

penguins_numeric <- map_lgl(penguins, is.numeric)
map_dbl(penguins[penguins_numeric], sd, na.rm = TRUE)
```

3. Compute the number of levels for every factor in the penguins dataset. Hint: you will use the technique from 2. to get numeric columns, and use length() in your function to get the number of levels.

```{r}
penguins_factor <- map_lgl(penguins, is.factor)
map_int(penguins[penguins_factor], ~ length(levels(.x)))
```

4. Using map_df(), create a report on the iris dataset summarizing each column. Include at least the column name, distinct n and type for each column.

```{r}
iris %>% map_df(~(data.frame(n_distinct = n_distinct(.x),
                class = class(.x))),
                .id = "variable"
)
```

5. Create another report on the iris dataset, giving the mean, max, min, and standard deviation for all the numeric variables. Use map_lgl() to get a logical vector of numeric columns, and then use map_df() to create the report.

```{r}
iris_num <- map_lgl(iris, is.numeric)
map_df(iris[iris_num], ~(data.frame(
  mean = mean(.x),
  max = max(.x),
  min = min(.x),
  sd = sd(.x)
)),
  .id = "variable")
```


6. Write a walk2() call to output gapminder data out to your project directory, split up by continent. Use write.csv in your walk2() function call.

```{r}
continents <- split(gapminder, gapminder$continent)

paths <- file.path(getwd(), paste0("continent-", names(continents), ".csv"))

walk2(continents, paths, write.csv)
```

7. Use map() to fit linear models (using lm() ) to the mtcars dataset using the formulas stored in this list:

formulas <- list(
  mpg ~ disp,
  mpg ~ I(1 / disp),
  mpg ~ disp + wt,
  mpg ~ I(1 / disp) + wt
)

```{r}
formulas <- list(
  mpg ~ disp,
  mpg ~ I(1 / disp),
  mpg ~ disp + wt,
  mpg ~ I(1 / disp) + wt
)

models <- map(formulas, lm, data = mtcars)
```

8. What does write.csv() return, i.e. what happens if you use it with map2() instead of walk2()?

```{r}
cyls <- split(mtcars, mtcars$cyl)
paths <- file.path(tempdir(), paste0("cyl-", names(cyls), ".csv"))

map2(cyls, paths, write.csv)
```

map2() returns NULL for each file.

