---
title: "R Open Lab 4 - Advanced ggplot2"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Advanced ggplot2

Today we will review the basics of data visualization in R using ggplot2, and then learn how to use new packages for other visualization types.

To learn more about visualizations in R, let's take a look at the R Graph Gallery:
https://r-graph-gallery.com

This site has lots of different plots you can make in R, with instructions on which package to use, and code examples to help you get started.

### Resources (crtl-click or cmd-click to open the pages)
This open lab was adapted from https://github.com/thomasp85/ggplot2_workshop (video for original workshop available at: https://www.youtube.com/watch?v=h29g21z0a68)
Visit https://go.ncsu.edu/getdatahelp for help with data science. You can email us, chat with a Data Science Consultant,
or book a 30-minute consultation with a Data Science Consultant.
To find out more about upcoming library workshops, visit https://lib.ncsu.edu/workshops.
The Data & Visualization workshops offered by DVS are at https://www.lib.ncsu.edu/workshops/category/data-and-visualization

### Eval: Please take a few minutes to fill out the workshop eval before you leave: https://go.ncsu.edu/dvs-eval

### Dependencies
This document comes with a list of required packages. 

```{r, echo=FALSE}
packages <- c(
  'ggplot2',
  'ggrepel',
  'patchwork',
  'ggraph',
  'dplyr',
  'gganimate',
  'gifski',
  'ggforce',
  'ggthemes',
  'maps',
  'sf',
  'tidyr',
  'concaveman',
  'remotes'
)
# Uncomment the next lines (remove the first #) to install the packages. Comment again when you have installed them.
install.packages("remotes")
install.packages(packages)
remotes::install_github("hrbrmstr/hrbrthemes")
```

```{r}
# Load standard packages
library(ggplot2)
```

### Datasets
We will use an assortment of datasets throughout the document. The purpose is 
mostly to showcase different plots, and less on getting some divine insight into
the world. While not necessary we will call `data(<dataset>)` before using a new
dataset to indicate the introduction of a new dataset.

## Review from Data Visualization with R
We will look at the basic ggplot2 use using the faithful dataset, giving 
information on the eruption pattern of the Old Faithful geyser in Yellowstone
National Park.

```{r}
data("faithful")
# Basic scatterplot
ggplot(data = faithful, 
       mapping = aes(x = eruptions, y = waiting)) + 
  geom_point()
# Data and mapping can be given both as global (in ggplot()) or per layer
ggplot() + 
  geom_point(mapping = aes(x = eruptions, y = waiting),
             data = faithful)
```

If an aesthetic is linked to data it is put into `aes()`

```{r}
ggplot(faithful) + 
  geom_point(aes(x = eruptions, y = waiting, colour = eruptions < 3))
```

If you simple want to set a value, put it outside of `aes()`

```{r}
ggplot(faithful) + 
  geom_point(aes(x = eruptions, y = waiting),
             colour = 'steelblue')
```

Some geoms only need a single mapping and will calculate the rest for you

```{r}
ggplot(faithful) + 
  geom_histogram(aes(x = eruptions))
```

geoms are drawn in the order they are added. The point layer is thus drawn on
top of the density contours in the example below.

```{r}
ggplot(faithful, aes(x = eruptions, y = waiting)) + 
  geom_density_2d() + 
  geom_point()
```

### Stat
We will use the `mpg` dataset giving information about fuel economy on different 
car models.

Every geom has a stat. This is why new data (`count`) can appear when using 
`geom_bar()`.

```{r}
data("mpg")
ggplot(mpg) + 
  geom_bar(aes(x = class))
```

The stat can be overwritten. If we have precomputed count we don't want any 
additional computations to perform and we use the `identity` stat to leave the 
data alone

```{r}
library(dplyr)
mpg_counted <- mpg %>% 
  count(class, name = 'count')
ggplot(mpg_counted) + 
  geom_bar(aes(x = class, y = count), stat = 'identity')
```

Most obvious geom+stat combinations have a dedicated geom constructor. The one
above is available directly as `geom_col()`

```{r}
ggplot(mpg_counted) + 
  geom_col(aes(x = class, y = count))
```

### Scales
Scales define how the mapping you specify inside `aes()` should happen. All 
mappings have an associated scale even if not specified.

```{r}
ggplot(mpg) + 
  geom_point(aes(x = displ, y = hwy, colour = class))
```

take control by adding one explicitly. All scales follow the same naming 
conventions.

Color brewer website: https://colorbrewer2.org/#type=sequential&scheme=BuGn&n=3
```{r}
ggplot(mpg) + 
  geom_point(aes(x = displ, y = hwy, colour = class)) + 
  scale_colour_brewer(type = 'qual')

# See our color palette options
RColorBrewer::display.brewer.all()

ggplot(mpg) + 
  geom_point(aes(x = displ, y = hwy, colour = class)) + 
  scale_colour_brewer(palette="Pastel1")
```

Positional mappings (x and y) also have associated scales.

```{r}
ggplot(mpg) + 
  geom_point(aes(x = displ, y = hwy)) + 
  scale_x_continuous(breaks = c(3, 5, 6)) + 
  scale_y_continuous(trans = 'log10')
```

### Facets
The facet defines how data is split among panels. The default facet 
(`facet_null()`) puts all the data in a single panel, while `facet_wrap()` and
`facet_grid()` allows you to specify different types of small multiples

```{r}
ggplot(mpg) + 
  geom_point(aes(x = displ, y = hwy)) + 
  facet_wrap(~ class)
```

```{r}
ggplot(mpg) + 
  geom_point(aes(x = displ, y = hwy)) + 
  facet_grid(year ~ drv)
```

### Theme
Theming defines the feel and look of your final visualisation and is something
you will normally defer to the final polishing of the plot. It is very easy to 
change looks with a prebuild theme

```{r}
ggplot(mpg) + 
  geom_bar(aes(y = class)) + 
  facet_wrap(~year) + 
  theme_minimal()
```

Further adjustments can be done in the end to get exactly the look you want

```{r}
ggplot(mpg) + 
  geom_bar(aes(y = class)) + 
  facet_wrap(~year) + 
  labs(title = "Number of car models per class",
       caption = "source: http://fueleconomy.gov",
       x = NULL,
       y = NULL) +
  scale_x_continuous(expand = c(0, NA)) + 
  theme_minimal() + 
  theme(
    text = element_text('Arial'),
    strip.text = element_text(face = 'bold', hjust = 0),
    plot.caption = element_text(face = 'italic'),
    panel.grid.major = element_line('white', size = 0.5),
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.ontop = TRUE
  )
```

## Extensions
While ggplot2 comes with a lot of batteries included, the extension ecosystem 
provides priceless additional features

### Animation
ggplot2 is usually focused on static plots, but gganimate extends the API and
grammar to describe animations. As such it feels like a very natural extension
of using ggplot2

```{r}
ggplot(economics) + 
  geom_line(aes(x = date, y = unemploy))
library(gganimate)
ggplot(economics) + 
  geom_line(aes(x = date, y = unemploy)) + 
  transition_reveal(along = date)
```

There are many different transitions that control how data is interpreted for
animation, as well as a range of other animation specific features

```{r}
anim <- ggplot(mpg) + 
  geom_bar(aes(x = factor(cyl)))
ggplot(mpg) + 
  geom_bar(aes(x = factor(cyl))) + 
  labs(title = 'Number of cars in {closest_state} by number of cylinders') + 
  transition_states(states = year) + 
  enter_grow() + 
  exit_fade()
anim_save("example.gif")
```

### Annotation
Text is a huge part of storytelling with your visualisation. Historically, 
textual annotations has not been the best part of ggplot2 but new extensions 
make up for that.

Standard geom_text will often result in overlaping labels
```{r}
ggplot(mtcars, aes(x = disp, y = mpg)) + 
  geom_point() + 
  geom_text(aes(label = row.names(mtcars)))
```

ggrepel takes care of that

```{r}
library(ggrepel)
ggplot(mtcars, aes(x = disp, y = mpg)) + 
  geom_point() + 
  geom_text_repel(aes(label = row.names(mtcars)))
```

If you want to highlight certain parts of your data and describe it, the 
`geom_mark_*()` family of geoms have your back

```{r}
library(ggforce)
ggplot(mtcars, aes(x = disp, y = mpg)) +
  geom_point() + 
  geom_mark_ellipse(aes(filter = gear == 4,
                        label = '4 gear cars',
                        description = 'Cars with fewer gears tend to both have higher yield and lower displacement'))
```

### Looks
Many people have already designed beautiful (and horrible) themes for you. Use
them as a base.

### Fonts and other resources
Note that you may need to install missing fonts and other theme resources. In the following example, my system is lacking the "Roboto" set of fonts. I had to add the font using showtext.

```{r warning=FALSE}
p <- ggplot(mtcars, aes(mpg, wt)) +
  geom_point(aes(color = factor(carb))) +
  labs(
    x = 'Fuel efficiency (mpg)', 
    y = 'Weight (tons)',
    title = 'Seminal ggplot2 example',
    subtitle = 'A plot to show off different themes',
    caption = 'Source: It’s mtcars — everyone uses it'
  )

library(hrbrthemes)

# Run this if you don't have showtext installed
# install.packages("showtext")
library(showtext)

# Add the Roboto font
font_add_google("Roboto Condensed", family="Roboto")
font_add_google("Roboto", family="Roboto")
showtext_auto()

p + theme_modern_rc()
# p + theme_economist()
# p + theme_fivethirtyeight()

p + 
   scale_colour_ipsum() + 
   theme_ipsum()
```

```{r}
library(ggthemes)
p + 
  scale_colour_excel() + 
  theme_excel()
```

### Eval: Please take a few minutes to fill out the workshop eval before you leave: https://go.ncsu.edu/dvs-eval
