---
title: "solutions"
author: "Joddy"
date: "2023-10-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Review
Modify the code below to make the points larger squares and slightly transparent.
See `?geom_point` for more information on the point layer.

```{r}
ggplot(faithful) + 
  geom_point(aes(x = eruptions, y = waiting), alpha=0.9, size=3, shape="square")
```

Hint 1: transparency is controlled with `alpha`, and shape with `shape`
Hint 2: remember the difference between mapping and setting aesthetics

See https://ggplot2.tidyverse.org/articles/ggplot2-specs.html for more details about alpha, shape, and size aesthetics.

* * *

Add a line that separates the two point distributions. See `?geom_abline` for 
how to draw straight lines from a slope and intercept.

```{r}
ggplot(faithful) + 
  geom_point(aes(x = eruptions, y = waiting)) +
  geom_vline(xintercept=3, color="red")
```

### Stat
While most people use `geom_*()` when adding layers, it is just as valid to add 
a `stat_*()` with an attached geom. Look at `geom_bar()` and figure out which
stat it uses as default. Then modify the code to use the stat directly instead
(i.e. adding `stat_*()` instead of `geom_bar()`)

```{r}
ggplot(mpg) + 
  stat_count(aes(x = class))
```

* * *

### Scales
Use `RColorBrewer::display.brewer.all()` to see all the different palettes from
Color Brewer and pick your favourite. Modify the code below to use it

```{r}
RColorBrewer::display.brewer.all()

ggplot(mpg) + 
  geom_point(aes(x = displ, y = hwy, colour = class)) + 
  scale_colour_brewer(palette="Pastel1")
```

* * *

### Facets
One of the great things about facets is that they share the axes between the 
different panels. Sometimes this is undesirable though, and the behavior can
be changed with the `scales` argument. Experiment with the different possible
settings in the plot below:

```{r}
ggplot(mpg) + 
  geom_point(aes(x = displ, y = hwy)) + 
  facet_wrap(~ drv, scales="free")

ggplot(mpg) + 
  geom_point(aes(x = displ, y = hwy)) + 
  facet_wrap(~ drv, scales="free_x")

ggplot(mpg) + 
  geom_point(aes(x = displ, y = hwy)) + 
  facet_wrap(~ drv, scales="free_y")
```

* * *

Facets can be based on multiple variables by adding them together. Try to 
recreate the same panels present in the plot below by using `facet_wrap()`

```{r}
ggplot(mpg) + 
  geom_point(aes(x = displ, y = hwy)) + 
  facet_grid(year ~ drv)

ggplot(mpg) + 
  geom_point(aes(x = displ, y = hwy)) + 
  facet_wrap(~ year + drv)
```

### Theme
Themes can be overwhelming, especially as you often try to optimize for beauty 
while you learn. To remove the last part of the equation, the exercise is to 
take the plot given below and make it as hideous as possible using the theme
function. Go absolutely crazy, but take note of the effect as you change 
different settings.

```{r}
ggplot(mpg) + 
  geom_bar(aes(y = class, fill = drv)) + 
  facet_wrap(~year) + 
  labs(title = "Number of car models per class",
       caption = "source: http://fueleconomy.gov",
       x = 'Number of cars',
       y = NULL)
```

### Animation
The animation below will animate between points showing cars with different 
cylinders.

```{r}
ggplot(mpg) + 
  geom_point(aes(x = displ, y = hwy, group=cyl)) + 
  ggtitle("Cars with {closest_state} cylinders") + 
  transition_states(factor(cyl))
```

gganimate uses the `group` aesthetic to match observations between states. By 
default the group aesthetic is set to the same value, so observations are 
matched by their position (first row of 4 cyl is matched to first row of 5 cyl 
etc.). This is clearly wrong here (why?). Add a mapping to the `group` aesthetic
to ensure that points do not move between the different states.

* * *

In the presence of discrete aesthetic mappings (`colour` below), the group is
deduced if not given. The default behavior of objects that appear and disappear
during the animation is to simply pop in and out of existence. `enter_*()` and
`exit_*()` functions can be used to control this behavior. Experiment with the
different enter and exit functions provided by gganimate below. What happens if 
you add multiple enter or exit functions to the same animation?

```{r}
ggplot(mpg) + 
  geom_point(aes(x = displ, y = hwy, colour = factor(cyl))) + 
  ggtitle("Cars with {closest_state} cylinders") + 
  transition_states(factor(cyl))
```

* * *

In the animation below (as in all the other animations) the changes happens at
constant speed. How values change during an animation is called easing and can
be controlled using the `ease_aes()` function. Read the documentation for 
`ease_aes()` and experiment with different easings in the animation.

```{r}
mpg2 <- tidyr::pivot_longer(mpg, c(cty,hwy))
ggplot(mpg2) + 
  geom_point(aes(x = displ, y = value)) + 
  ggtitle("{if (closest_state == 'cty') 'Efficiency in city' else 'Efficiency on highway'}") + 
  transition_states(name) +
  ease_aes(y="bounce-in")

mpg2 <- tidyr::pivot_longer(mpg, c(cty,hwy))
ggplot(mpg2) + 
  geom_point(aes(x = displ, y = value)) + 
  ggtitle("{if (closest_state == 'cty') 'Efficiency in city' else 'Efficiency on highway'}") + 
  transition_states(name) +
  ease_aes(y="quadratic-out")

mpg2 <- tidyr::pivot_longer(mpg, c(cty,hwy))
ggplot(mpg2) + 
  geom_point(aes(x = displ, y = value)) + 
  ggtitle("{if (closest_state == 'cty') 'Efficiency in city' else 'Efficiency on highway'}") + 
  transition_states(name) +
  ease_aes(y="elastic-in-out")
```

### Annotation
Explore the documentation for `geom_text_repel`. Find a way to ensure that the 
labels in the plot below only repels in the vertical direction

```{r}
mtcars$label <- ""
mtcars$label[1:10] <- rownames(mtcars)[1:10]
ggplot(mtcars, aes(x = disp, y = mpg)) + 
  geom_point() + 
  geom_text_repel(aes(label = label), direction="y", nudge_y=10)
```

* * *

ggforce comes with 4 different types of mark geoms. Try them all out in the code
below:

```{r}
ggplot(mtcars, aes(x = disp, y = mpg)) +
  geom_point() + 
  geom_mark_ellipse(aes(filter = gear == 4,
                        label = '4 gear cars'))

ggplot(mtcars, aes(x = disp, y = mpg)) +
  geom_point() + 
  geom_mark_circle(aes(filter = gear == 4,
                        label = '4 gear cars'))

ggplot(mtcars, aes(x = disp, y = mpg)) +
  geom_point() + 
  geom_mark_rect(aes(filter = gear == 4,
                        label = '4 gear cars'))

ggplot(mtcars, aes(x = disp, y = mpg)) +
  geom_point() + 
  geom_mark_hull(aes(filter = gear == 4,
                        label = '4 gear cars'))
```